---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Sehar Bajwa
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
prefer-html: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
#install.packages("usethis")
#install.packages("gitcreds")
#library(usethis)
#library(gitcreds)
#create_github_token()
# Example dataframe with values and ranges
data_ranges <- data.frame(
  category = c("A", "B", "C", "D"),
  value = c(100, 150, 200, 250),
  range = c(10, 15, 20, 25)  # Example range values
)

# Calculate upper and lower bounds for plotting
data_ranges$lower <- data_ranges$value - data_ranges$range
data_ranges$upper <- data_ranges$value + data_ranges$range

# Print the dataframe
print("Data with Ranges:")
print(data_ranges)
# Load necessary library
library(ggplot2)

# Create the plot
ggplot(data_ranges, aes(x = category, y = value)) +
  geom_point() +  # Plot points for values
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +  # Error bars for ranges
  labs(title = "Values with Ranges", x = "Category", y = "Value") +
  theme_minimal()




library(tidyverse)
library(palmerpenguins)
```


# Introduction

In the biggest democratic election the world has ever seen, India held its general election in 2024 over 7 successive phases, beginning on the 19th of April and ending on the 1st of June. The election determined the composition of the Lok Sabha, the lower house of India's Parliament, which comprises 543 seats. A party or coalition needs a minimum of 272 seats to form a government.

During the election period, numerous exit and opinion polls were conducted to gauge public sentiment and predict the potential outcome. Opinion polls were conducted in the months leading up to the election, providing insights into voter preferences and trends. Exit polls were conducted immediately after voters cast their ballots, offering a snapshot of the electorate's choices.

The results of these polls were keenly watched by political analysts, parties, and the public, as they provided an early indication of which party or coalition might secure a majority in the Lok Sabha. Despite their limitations and varying degrees of accuracy, these polls played a crucial role in shaping the narrative and expectations surrounding the 2024 Indian general election.

# Background  {#sec-background}

Before delving into the following sections of the paper, it is important to provide some political context. 
The National Democratic Alliance (NDA), led by the Bharatiya Janata Party (BJP), has been a significant political force in India since its formation in the late 1990s, particularly under Prime Minister Narendra Modi, who assumed office in 2014 and was re-elected in 2019.
In 2023, the Indian National Developmental Inclusive Alliance (INDIA) was established, comprising major opposition parties from across the political spectrum, including the Indian National Congress (INC), the Dravida Munnetra Kazhagam (DMK), and the Trinamool Congress (TMC). This coalition, headed by Rahul Gandhi of the INC, was formed to present a united front against the BJP's dominance, driven by concerns over Modi's prolonged tenure. The INDIA Alliance's collaboration aimed to make a dent in the NDA's mammoth vote bank by consolidating votes and pooling in resources. And they did achieve the goal of inhibiting the NDA's expected landslide victory (353 seats in 2019). The slogans of '400 paar' (crossing 400 seats) by the NDA this time around were paid a lot of heed, and INDIA's alliance was underestimated by the opinion polls, and perhaps falsely represented by the exit polls.

The newshouses received a lot of flak for the inflation of the exit poll numbers in particular, nfluenced by the BJP's past electoral success and media narrative. In addition to misleading the public, they led to the worst stock market crash in the country in the past 4 years.The term "godi media" colloquially describes Indian media outlets perceived as biased towards the BJP, often criticized for their lack of independent journalism.

Consequently, the analysis in this paper will focus on the National Democratic Alliance (NDA) because of the central role it played in the 2024 Indian general election. With Prime Minister Narendra Modi seeking a third consecutive term, the election was largely seen as a referendum on his leadership and the performance of his Bharatiya Janata Party (BJP). The political discourse and public sentiment were heavily polarized around the NDA, with debates and predictions revolving around whether Modi and the BJP could maintain their stronghold on Indian politics. Consequently, all bets and polls were discussions around the electoral landscape being pro or anti-NDA, instead of being pro INDIA or pro Independent candidates.


# Data {#sec-data}

The data used in this paper is obtained by scraping the Wikipedia page for 'general elections in India 2024', and cross-referenced with each article it mentioned. There seems to be no singular database to collect all pre-election and exit poll data, and this is part of the gaping problem examined in this paper. The data points are determined from several National and local newspapers that have published the news online.

While nothing except the results themselves, possible margins, and the sample sizes for the opinion polls are available online, and when one follows links to the actual research/polling agency hired by the news outlets, there are no publicly available reports available on their websites either. 

The three data sets I will be using are Opinion Poll data, Exit Poll data and Mood of the Nation (CVoter).


## Opinion Poll data 

In the run-up to the 2024 General Election, various newspapers and agencies conduct pre-election polls to gauge the mindset, mood, and political leaning of the average voter. These polls provide valuable insights into public opinion, allowing political analysts, parties, and the general public to understand potential election outcomes better. They help identify trends, shifts in voter preferences, and the impact of current events and policies on voter behavior.

Containing data from January 2023 up to April 2024, this data consists of the following 7 variables: name of the polling agency, sample size, seats won by NDA, INDIA, and Others. Additionally, the 'Exact' date published is a self-created variable after scouring the references, so a more accurate dataset can be made, and the overlap in a line graph can be avoided. A snapshot of the cleaned data is provided below, which includes the 7 variables and 17 data points in total. 



```{r}
#| include: false
#| warning: false
#| message: false

#### Initialize quarto workspace by reading csv's ####
library(tidyverse)
library(here)
library(readr)
library(ggplot2)

cleaned_opinion_polldata <- read_csv(here::here("/cloud/project/data/analysis_data/cleaned_opinion_poll.csv"))
cleaned_exit_polldata <- read_csv(here::here("/cloud/project/data/analysis_data/cleaned_exit_poll.csv"))
MOTN_polldata <- read_csv(here::here("/cloud/project/data/raw_data/MOTN.csv"))


```

```{r}
#| echo: false
#| warning: false
#| label: tbl-dataa
#| tbl-cap: "A summary atable of cleaned data"
#install.packages("knitr")
#install.packages("kableExtra")
library(kableExtra)
library(knitr)
kable(cleaned_opinion_polldata, padding = 2)%>%
  kable_styling(font_size = 8) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "2cm") %>%
  column_spec(3, width = "2cm") %>%
  column_spec(4, width = "2cm") %>%
  column_spec(5, width = "1.5cm") %>%
  column_spec(6, width = "1.5cm") %>%
  column_spec(7, width = "1.5cm")

```

## Mood of the Nation data

Conducted by National Newshouse India Today, the CVoter opinion poll utilizes Computer Assisted Telephone Interviewing among voters aged 18 and above across all 543 Lok Sabha seats. Targeting a sample of 30,000 per quarter with an average response rate of 55%, the tracker is conducted weekly in 11 national languages. According to the official site, CVoter employs a Standard Resilient Distributed dataset to randomly select numbers across telecom lines, ensuring a representative analysis through statistical weighting of the data to reflect the local population as per the latest Census figures. The data is weighted to align with known census profiles, including gender, age, education, income, religion, caste, urban/rural distribution, and past vote recall for the Lok Sabha and Vidhan Sabha elections. For analytics, CVoter uses a proprietary algorithm to calculate provincial and regional vote shares based on the split-voter phenomenon.

The CVoter data offers the most consistent output for opinion polls in India and has been selected due to its standardization and regularity. This consistency allows for the assumption that all other variables are controlled, enabling a focused analysis of trends. The dataset has been independently constructed by aggregating infographics and articles published on the India Today website. The included dataset only has 5 variables, the month, and NDA, INDIA and Others Seats. A snapshot is provided below.

```{r}
#| echo: false
#| warning: false
#| label: tbl-dDAata
#| tbl-cap: "A summary table of cleaned data"
#install.packages("knitr")
#install.packages("kableExtra")
library(kableExtra)
library(knitr)
kable(MOTN_polldata, padding = 2)%>%
  kable_styling(font_size = 8) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "2cm") %>%
  column_spec(3, width = "2cm") %>%
  column_spec(4, width = "2cm") 


```

## Exit poll data

The inception of exit polls in India traces back to 1957, during the second Lok Sabha elections, with the pioneering initiative led by the Indian Institute of Public Opinion.

Exit polls, as the name suggests, are conducted after the culmination of the last phase of voting, in accordance with the guidelines set by the Election Commission of India. Distinguished from opinion polls, which precede elections, exit polls involve surveys administered to voters as they exit polling stations post casting their ballots. These surveys encompass a spectrum of inquiries, delving into voter motivations and party preferences.. As the 2024 Lok Sabha elections concluded, exit poll results for 28 states and eight union territories were slated for release from 6:30 pm onwards on June 1 2024, in compliance with Election Commission protocols.

In contrast to opinion polls, exit polls are often conducted and released hastily on the same day, rendering the timing irrelevant. Moreover, these exit polls have not disclosed the sample size of the respondents surveyed. As a result, they are presented here merely as indicators, highlighting their potential for inaccuracy and inflation. There are 4 variables, the name of the agency, and the number of seats predicted each for NDA, INDIA and others. The 15 datapoints have been scraped from Wikipedia using rvest. 

```{r}
#| echo: false
#| warning: false
#| label: tbl-data
#| tbl-cap: "A summary table of cleaned data"
#install.packages("knitr")
#install.packages("kableExtra")
library(kableExtra)
library(knitr)
kable(cleaned_exit_polldata, padding = 2)%>%
  kable_styling(font_size = 8) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "2cm") %>%
  column_spec(3, width = "2cm") %>%
  column_spec(4, width = "2cm") 

```





\newpage


# References


